---
title: "data.table"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
---

```{r setup, include=FALSE}
library(flexdashboard)
```

Getting Started
===================================== 
Column {data-width=700}
-----------------------------------------------------------------------

### Packages and data
```{r start, echo = TRUE}
# Packages: -------------------------------------------------------------------
library(tidyverse)
library(dbplyr)
library(Lahman)
library(data.table)
#lapply( c('tidyverse', 'dbplyr', 'Lahman', 'data.table'), install.packages)

# Create a local SQLlite database of the Lahman data: ----------------------
lahman = lahman_sqlite()

# Copy the batting table to memory as a tibble: -------------------------------
batting_tbl = lahman %>% 
  tbl("BATTING") %>% 
  collect()
class(batting_tbl)

# Convert the copy in memory to a data.table: ---------------------------------
batting_dt = as.data.table(batting_tbl)
class(batting_dt)
```

Column {data-width=300}
-----------------------------------------------------------------------
### Vignettes

  + [Introduction to data.table](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html)

  + [Reference Semantics](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reference-semantics.html)

  + [Efficient reshaping using data.tables](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-reshape.html)
  
  + The following points from the [datatable-faq](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-faq.html):
     1.5, 1.6, 2.1, 2.3, 2.16, 2.17, 2.21, 2.23

Select with `j`
===================================== 
Column {data-width=600}
-----------------------------------------------------------------------

### data.table
```{r j1_dt, echo = TRUE}
# data.table
batting_dt[ , .(playerID, yearID, league = lgID, stint)]  
```

The `j` clause in a data.table is used to select columns. The `.()` is an
alias for `list()` within a data.table. Each element of the list becomes a 
column in the resulting data.table. We can provide new names for the columns
by naming the list elements. 

Column {.tabset data-width=400}
-----------------------------------------------------------------------

### SQL
```{r j1_sql, echo = TRUE}
#SQL: SELECT
query = 
'
SELECT playerID, yearID, lgID as league, stint
FROM BATTING
'
lahman %>% tbl(sql(query)) 
```

### dplyr
```{r j1_dplyr, echo = TRUE }
# dplyr: select, rename, transmute
batting_tbl %>%
    transmute(playerID, yearID, league = lgID, stint)

batting_tbl %>%
    select(playerID, yearID, league = lgID, stint)
```

Compute in `j`
===================================== 
Column {.tabset data-width=600}
-----------------------------------------------------------------------

### data.table
```{r j2_dt, echo = TRUE}
# data.table
batting_dt[ , .(playerID, avg = H / AB)]
```

Any valid R expression can be used in "j" and we may reference the columns as
variables. This is similar to the behavior of the R function `with()`. As the 
long as the expression returns a *list* the result is a data.table. 

Column {.tabset data-width=400}
-----------------------------------------------------------------------

### SQL
```{r j2_sql, echo = TRUE}
#SQL: 
query = 
'
SELECT playerID, 
       (Cast (H as Float) /  Cast (AB as Float) ) as avg
FROM BATTING
'
lahman %>% tbl(sql(query)) %>% collect()
```

### dplyr
```{r j2_dplyr, echo = TRUE }
# dplyr: transmute
batting_tbl %>%
  transmute( playerID, avg = H / AB)
```

Aggregate in `j`
===================================== 
Column {data-width=600}
-----------------------------------------------------------------------

### data.table
```{r j3_dt, echo = TRUE}
# data.table: aggregate using a valid R expression
batting_dt[, .(max_HBP = max(HBP, na.rm=TRUE))]
```

Here we find the maximum HBP entry in the table using a function that returns
a single value. 

Column {.tabset data-width=400}
-----------------------------------------------------------------------

### SQL
```{r j3_sql, echo = TRUE}
# SQL: SELECT with aggregation function
query =
'
SELECT max(HBP) as max_hbp
FROM BATTING
'
lahman %>% tbl(sql(query)) %>% collect()
```

### dplyr
```{r j3_dplyr, echo = TRUE }
# dplyr: summarize
batting_tbl %>%
  summarize( max_HBP = max(HBP, na.rm=TRUE))
```

Grouping with `by`
===================================== 
Column {.tabset data-width=600}
-----------------------------------------------------------------------

### data.table: `by`
```{r j4_dt, echo = TRUE}
# data.table
#batting_dt[ , .(avg = sum(H) / sum(AB)), by = .(playerID, yearID, lgID)]
batting_dt[ , .(avg = sum(H) / sum(AB)), .(playerID, yearID, lgID)]
```

We can perform grouped operations in `j` using the `by` argument. Use a list
to form groups using the unique combinations from multiple columns. Observe that
unlike the SQL and dplyr versions the resulting table is not re-ordered by 
group.

### data.table: `keyby`
```{r j4_dt_keyby, echo = TRUE}
# data.table: keyby
# Here the parameter name is required
batting_dt[ , .(avg = sum(H) / sum(AB)), keyby = .(playerID, yearID, lgID)]
```

If we use `keyby` the new data.table is reordered by group.

Column {.tabset data-width=400}
-----------------------------------------------------------------------

### SQL
```{r j4_sql, echo = TRUE}
#SQL: 
query = 
'
SELECT playerID, yearID, lgID,
       ( Cast (sum(H) as Float) / Cast (sum(AB) as Float) ) as avg
FROM BATTING
GROUP BY playerID, yearID, lgID
'
lahman %>% tbl(sql(query)) %>% collect()
```

### dplyr
```{r j4_dplyr, echo = TRUE }
# dplyr: group_by
batting_tbl %>% 
  group_by(playerID, yearID, lgID) %>%
  summarize( avg = sum(H) / sum(AB) )
```

Select rows in `i`
===================================== 
Column {.tabset data-width=600}
-----------------------------------------------------------------------

### data.table: logical indexing
```{r j5_dt, echo = TRUE}
# data.table
batting_dt[ yearID == 2016, .(playerID, HBP)]
```

The `i` clause in a data.table is used to select rows. Similar to `dplyr::filter`
or subsetting a data.frame we often use it with a logical index whose length matches
the number of rows. 

### data.table: keys
```{r j5_dt_key, echo = TRUE}
# data.table
setkey(batting_dt, 'teamID')
batting_dt['DET', .(playerID, teamID, HBP)]
key(batting_dt)

setkey(batting_dt, 'yearID')
batting_dt[.(2016), .(playerID, yearID, HBP)]

# Compare the difference
batting_dt[2016, .(playerID, yearID, HBP)]

# Remove key for later examples
setkey(batting_dt, NULL)
```

If our data.table has one or more columns designated as a key, we can subset by
value in `i`.  Note the need to pass integer-valued keys within a list to 
differentiate from simple subsetting by integer. 

When we set a key using `setkey()` the rows are reordered to enable faster
subsetting by key. This is similar to "indexing" in SQL. Read more about keys in
this [vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-keys-fast-subset.html). 

Column {.tabset data-width=400}
-----------------------------------------------------------------------

### SQL
```{r j5_sql_a, echo = TRUE}
#SQL: WHERE
query = 
'
SELECT playerID, HBP
FROM BATTING
WHERE yearID = 2016
'
lahman %>% tbl(sql(query)) %>% collect()
```

### dplyr
```{r j5_dplyr, echo = TRUE }
# dplyr: filter
batting_tbl %>%
  filter(yearID == 2016) %>%
  select(playerID, HBP)
```

### base R
```{r echo = TRUE}
batting_tbl[ batting_tbl$yearID == 2016, c("playerID", "yearID", "HBP")]
```

Chaining
===================================== 
Column {.tabset data-width=600}
-----------------------------------------------------------------------

### data.table: pipes
```{r, echo = TRUE}
# data.table
batting_dt[ yearID > 2000, .(HR = sum(HR)), .(playerID)] %>%
  .[HR > 400]
```

Those familiar with `dplyr` will be familiar with the "pipe" operation `%>%` that
calls the function on the right with the object on the left (usually data) as its
first parameter.  We can take advantage of the `.` used by `margittr::%>%` to use
pipes with data.table's `[]` function. 

### data.table: chaining

```{r}
batting_dt[ yearID > 2000, .(HR = sum(HR)), .(playerID)][HR > 400]
```

You can also simply "chain" the `[]` statements together. 

Column {.tabset data-width=400}
-----------------------------------------------------------------------

### SQL
```{r chaining, echo = TRUE}
#SQL: nested anonymous table, "HAVING"
query = 
'
SELECT *
FROM (
 SELECT playerID, sum(HR) as HR
 FROM BATTING
 WHERE yearID > 2000
 GROUP BY playerID
) 
WHERE HR > 400
'
```

### dplyr
```{r dplyr_pipe, echo = TRUE }
# dplyr: Use %>% to chain
batting_tbl %>% 
  filter(yearID > 2000) %>%
  group_by(playerID) %>%
  summarize( HR = sum(HR) ) %>%
  ## Here's the pipe from above
  filter( HR > 400)
```


Order in `i`
===================================== 
Column {data-width=600}
-----------------------------------------------------------------------

### data.table
```{r, echo = TRUE}
# data.table
batting_dt[ yearID > 2000, .(HR = sum(HR)), .(playerID)][HR > 400][order(-HR)]
```

Use `order()` in the `i` clause to sort a data.table. 


#### From `help(data.table::order)`:

"Also x[order(.)] is now optimised internally to use data.table's fast order. data.table always reorders in "C-locale" (see Details). To sort by session locale, use x[base::order(.)]."

Column {.tabset data-width=400}
-----------------------------------------------------------------------

### SQL
```{r sql_order, echo = TRUE}
#SQL: 
query = 
'
SELECT *
FROM (
 SELECT playerID, sum(HR) as HR
 FROM BATTING
 WHERE yearID > 2000
 GROUP BY playerID
) 
WHERE HR > 400
ORDER BY -HR
'
lahman %>% tbl(sql(query))  %>% collect()
```

### dplyr
```{r dplyr_pipes, echo = TRUE }
# dplyr: arrange
batting_tbl %>% 
  filter(yearID > 2000) %>%
  group_by(playerID) %>%
  summarize( HR = sum(HR) ) %>%
  ## Here's the pipe from above
  filter( HR > 400) %>%
  arrange( desc(HR) )
```

Using `.N`
===================================== 
Column {.tabset data-width=600}
-----------------------------------------------------------------------

### data.table: multiple stints
```{r, echo = TRUE}
# data.table: stints example
batting_dt[yearID == 2016 & AB > 99, .N, .(playerID)][N>1]
```

In SQL we often use COUNT to determine *how many* records in a group meet some
condition. In dplyr we can use `n()` within a function to count the number of
rows by group. In `data.table` the special symbol `.N` plays the same role. 

Here we use `.N` to find how many players have more than 100 AB with two or more teams in 2016. 

### data.table: 20-20
```{r, echo = TRUE}
# data.table: 20-20
batting_dt[yearID > 2000 , .(SB = sum(SB), HR = sum(HR)), .(playerID, yearID)] %>%
  .[SB > 19 & HR > 19] %>% .[ , .N, yearID]
```

Here we use `.N` to count the number of "20-20" players in each year since 2000. 

Column {.tabset data-width=400}
-----------------------------------------------------------------------

### SQL: stints
```{r sql_stints, echo = TRUE}
#SQL: stints example
query = 
'
SELECT playerID, COUNT(teamID) as N
FROM BATTING
WHERE yearID = 2016 AND AB > 99
GROUP BY playerID
HAVING N > 1
'
lahman %>% tbl(sql(query)) %>% collect()
```

### SQL: 20-20

```{r sql_20, echo = TRUE}
#SQL: 20-20
query = 
'
SELECT yearID, COUNT(playerID) as N
FROM (
 SELECT playerID, yearID, sum(SB) as SB, sum(HR) as HR
 FROM BATTING
 WHERE yearID > 2000
 GROUP BY playerID, yearID
 HAVING SB > 19 AND HR > 19
)
GROUP BY yearID
'
lahman %>% tbl(sql(query)) %>% collect()
```

### dplyr: stints
```{r dplyr_stints, echo = TRUE }
# dplyr: n()
batting_tbl %>% 
  filter(AB > 99 & yearID == 2016) %>%
  group_by(playerID, yearID) %>%
  summarize(N=n()) %>%
  filter(N > 1)
```

### dplyr: 20-20
```{r dplyr_20, echo = TRUE}
# dplyr: n()
batting_tbl %>%
  filter( yearID > 2000) %>%
  group_by(yearID, playerID) %>%
  summarize( SB = sum(SB), HR = sum(HR) ) %>%
  filter( SB > 19 & HR > 19) %>%
  summarize( n = n() )
```

Exercises 
===================================== 
Column {.tabset data-width=500}
-----------------------------------------------------------------------

### data.table: Instructions

Write data.table code to match each of the SQL or dplyr queries to the right.

1. Find the 2016 average for all playerID with team "DET" and at least 100 AB.

2. Find the playerID from team "DET" with the highest average in 2016, with
a minimum of 100 AB.

3. Find the playerID who hit the most total HRs between 2001 and 2010.

4. Find all playerIDs with at least 200 hits in 2016 (across all stints) sorted in
descending order.

5. Find the number of playerIDs with at least 200 hits in each year since 2000. 

### solution-1
```{r ex_dt1, echo = TRUE}
# data.table
batting_dt[ teamID == 'DET' & yearID == 2016 & AB >= 100, 
            .(playerID, avg = H / AB)]
```

### solution-2
```{r ex_dt2, echo = TRUE}
# data.table
batting_dt[ teamID == 'DET' & yearID == 2016 & AB >= 100, 
            .(playerID, avg = H / AB)][avg==max(avg)]
```

### solution-3
```{r ex_dt3, echo = TRUE}
# data.table
batting_dt[yearID %in% 2001:2010, .(HR = sum(HR)), .(playerID)] %>%
 .[which.max(HR)]
```

### solution-4
```{r ex_dt4, echo = TRUE}
# data.table
batting_dt[yearID==2016, .(H = sum(H)), .(playerID) ][ H>=200 ][ order(-H) ]
```

### solution-5
```{r ex_dt5, echo = TRUE}
# data.table
batting_dt[yearID > 1999, .(H = sum(H)), .(playerID, yearID) ][ H>=200 ] %>%
   .[ , .N, yearID] %>% .[order(-yearID)]
```

Column {.tabset data-width=500}
-----------------------------------------------------------------------

### SQL-1
```{r ex_sql1, echo = TRUE}
query = 
'
SELECT playerID, ( Cast (H as Float) / Cast (AB as Float) ) as avg
FROM BATTING 
WHERE teamID = "DET" AND yearID = 2016 AND AB > 99
'
lahman %>% tbl(sql(query)) %>% collect()
```

### SQL-2
```{r ex_sql2, echo = TRUE}
query = 
'
SELECT playerID, max( Cast (H as Float) / Cast (AB as Float) ) as avg
FROM BATTING 
WHERE teamID = "DET" AND yearID = 2016 AND AB > 99
'
lahman %>% tbl(sql(query)) %>% collect()
```

### SQL-3
```{r ex_sql3, echo = TRUE}
query = 
'
SELECT playerID, max(HR)
FROM (
 SELECT playerID, sum(HR) as HR
 FROM BATTING 
 WHERE yearID < 2011 AND yearID > 2000
 GROUP BY playerID
) 
'
lahman %>% tbl(sql(query)) %>% collect()
```

### SQL-4
```{r ex_sql4, echo = TRUE}
query = 
'
SELECT playerID, sum(H) as H
FROM BATTING 
WHERE yearID = 2016
GROUP BY playerID
HAVING H > 199
ORDER BY -H
'
lahman %>% tbl(sql(query)) %>% collect()
```

### SQL-5
```{r ex_sql5, echo = TRUE}
query = 
'
SELECT yearID as year, COUNT(playerID) as N
FROM (
 SELECT playerID, yearID, sum(H) as H
 FROM BATTING 
 WHERE yearID > 1999
 GROUP BY playerID, yearID
 HAVING H > 199
) 
GROUP BY yearID
ORDER BY -yearID
'
lahman %>% tbl(sql(query)) %>% collect()
```

### dplyr-1 
```{r ex_dplyr1, echo = TRUE }
batting_tbl %>%
  filter(yearID == 2016 & teamID == 'DET' & AB > 100) %>%
  transmute(playerID, avg = H / AB)
```

### dplyr-2 
```{r ex_dplyr2, echo = TRUE }
batting_tbl %>%
  filter(yearID == 2016 & teamID == 'DET' & AB > 100) %>%
  transmute(playerID, avg = H / AB) %>%
  filter(avg == max(avg))
```

### dplyr-3 
```{r ex_dplyr3, echo = TRUE }
batting_tbl %>%
  filter(yearID %in% 2001:2010) %>%
  group_by(playerID) %>%
  summarize(HR = sum(HR)) %>%
  filter(HR == max(HR))
```

### dplyr-4 
```{r ex_dplyr4, echo = TRUE }
batting_tbl %>%
  filter(yearID %in% 2016) %>%
  group_by(playerID) %>%
  summarize(H = sum(H)) %>%
  filter(H > 199) %>%
  arrange( desc(H) )
```

### dplyr-5 
```{r ex_dplyr5, echo = TRUE }
batting_tbl %>%
  filter(yearID > 1999) %>%
  group_by(yearID, playerID) %>%
  summarize(H = sum(H)) %>% # still grouped by yearIDs
  filter(H > 199) %>%       # or just summarize( n = sum(H > 199))
  summarize( n = n() ) %>%
  arrange( desc(yearID) )
```


